Avoiding tipping points in fisheries management: A Bayesian Non-Parametric approach

### 2014-05-29 manuscripts/manuscript.Rmd
new sensitivity analysis

keywords:
  - Bayesian
  - Structural Uncertainty
  - Nonparametric 
  - Optimal Control
  - Decision Theory
  - Gaussian Processes
  - Fisheries Management
---

@@

```{r export-data}
write.csv(sensitivity_dat, "sensitivity.csv")
```

@@

@@

```{r figure_2, dependson=c("assemble-models", "plot-options"), fig.cap="Graph of the inferred Gaussian process compared to the true process and maximum-likelihood estimated process.  Graph shows the expected value for the function $f$ under each model.  Two standard deviations from the estimated Gaussian process covariance with (light grey) and without (darker grey) measurement error are also shown.  The training data is also shown as black points.  The GP is conditioned on (0,0), shown as a pseudo-data point.", fig.width=8, fig.height=6}
x_grid_short <- x_grid[1:40]
gp_short <- gp_predict(gp, x_grid_short, burnin=1e4, thin=300)
models_posteriors <- 
  melt(lapply(sample(100, 50), 
              function(i){
    sample_gp <- mvrnorm(1, 
                            gp_short$Ef_posterior[,i],         
                            gp_short$Cf_posterior[[i]])
    data.frame(stock = x_grid_short, 
               GP = sample_gp,
               y = sample_gp,
               ymin = sample_gp - 2 * sqrt(gp_short$E_Vf), 
               ymax = sample_gp + 2 * sqrt(gp_short$E_Vf), 
               True = sapply(x_grid_short,f,0, p),  
               MLE = sapply(x_grid_short,f,0, est$p), 
               Allen = sapply(x_grid_short, allen_f, 0, pardist[i,]), 
               Ricker = sapply(x_grid_short, ricker_f, 0, ricker_pardist[i,]), 
               Myers = sapply(x_grid_short, myers_f, 0, myers_pardist[i,]))
             }), 
       id=c("stock", "y", "ymin", "ymax"))
ggplot(models_posteriors) + 
    geom_ribbon(aes(x=stock, y=y, ymin=ymin, ymax=ymax, group=L1), 
                  fill = "gray80", 
                  data=subset(models_posteriors, variable == "GP")) + 
    geom_line(aes(stock, value, col = variable, 
                  group=interaction(L1,variable)), 
              alpha=.2) + 
    geom_point(data = obs, aes(x,y), alpha = 0.8) + 
    xlab(expression(X[t])) + ylab(expression(X[t+1])) +
    facet_wrap(~variable) + 
    scale_colour_manual(values=colorkey) +  
    theme(legend.position="none")
```

@@

```{r figure_3, fig.width=8, fig.height=6, fig.cap="Outside the range of the training data (Figure 1), the true dynamics (black dots) fall outside the uncertainty (two standard deviations, colored bands) of the structurally incorrect parametric models (Ricker, Myers), but inside the uncertainty predicted by the GP. Points show the stock size simulated by the true model.  Overlay shows the range of states predicted by each model, based on the state observed in the previous time step. The Ricker model always (wrongly) predicts positive population growth, while the actual population shrinks in each step as the initial condition falls below the Allee threshold of the underlying model (Allen).  Note that because it does not assume a parametric form but instead relies more directly on the data, the GP is both more pessimistic and more uncertain about the future state than the parametric models.", dependson="plot-options"}
y <- numeric(8)
y[1] <- 4.5
for(t in 1:(length(y)-1))
      y[t+1] = z_g() * f(y[t], h=0, p=p)
# predicts means, does not reflect uncertainty estimate!
crash_data <- step_ahead_posteriors(y)
crash_data <- subset(crash_data, variable %in% c("GP", "Allen", "Ricker", "Myers"))
ggplot(crash_data) + 
  geom_boxplot(aes(as.factor(as.integer(time)), value, 
                   fill = variable, col=variable), 
               alpha=.7, outlier.size=1, position="identity") + 
#  geom_line(aes(time, value, col = variable, 
#            group=interaction(L1,variable)), alpha=.1) + 
  geom_point(aes(time, stock), size = 3) + 
  scale_fill_manual(values=colorkey[c("GP", "Allen", "Ricker", "Myers")], 
                      guide = guide_legend(override.aes = list(alpha = 1))) +  
  scale_colour_manual(values=colorkey[c("GP", "Allen", "Ricker", "Myers")], 
                      guide = guide_legend(override.aes = list(alpha = 1))) +  
  facet_wrap(~variable) + 
  theme(legend.position="none") + xlab("time") + ylab("stock size") 
```

@@

ggplot(policies, aes(stock, stock - value, color=method)) +
  geom_line(lwd=1.2, alpha=0.8) + xlab("stock size") + ylab("escapement")  +
  scale_colour_manual(values=colorkey)
```

@@

@@

# ggplot(actual_over_optimal, aes(value)) + geom_histogram(aes(fill=variable), binwidth=0.1) + 
#  xlab("Total profit by replicate")+ scale_fill_manual(values=colorkey)
# ggplot(actual_over_optimal, aes(value, fill=variable, color=variable)) + # density plots fail when delta fn
#  stat_density(aes(y=..density..), position="stack", adjust=3, alpha=.9) + 
#  xlab("Total profit by replicate")+ scale_fill_manual(values=colorkey)+ scale_color_manual(values=colorkey)
```

@@

(following the Ahrenius relationship, see @Gardiner2009 for the derivation).
As a result, the impact of using a model which underestimates the risk of harvesting
past the critical point is much more severe, since this such a situation
occurs more often.  Conversely, with large enough distance between the
optimal escapement and unstable points relative to the noise, the chance
of a transition becomes vanishingly small and all models can be modeled
near-optimally. Models that underestimate the cost incurred by stock
sizes fluctuating significantly below the optimal escapement level will
not perform poorly as long as those fluctuations are sufficently rare.

@@

```{r figure_7, fig.height=6, dependson=c("sensitivity-calc", "export-data"), fig.cap="Sensitivity Analysis.  Histogram shows the ratio of the realized net present value derived when managing under the GPDP over the optimal value given the true model and true parameters. Values near 1 indicate nearly optimal performance. Simulations cover a range of underlying parameter, model, replicate, and noise levels."}
ggplot(sensitivity_dat) + 
  geom_histogram(aes(value, fill=noise)) + 
  xlim(0,1.0) + 
  theme_bw() + 
  xlab("value as fraction of the optimal") + 
  facet_wrap(~model, ncol=1)
```

@@

```{r figure_S4, fig.cap="Posteriors from the MCMC estimate of the Ricker model", fig.width=6, fig.height=4}
ggplot(ricker_posteriors, aes(value)) + 
  stat_density(geom="path", position="identity", alpha=0.7) +
  facet_wrap(~ variable, scale="free", ncol=2)
```

@@

```{r figure_S6, fig.cap="Posterior distributions from the MCMC estimates of the Myers model", fig.width=6, fig.height=6}
ggplot(myers_posteriors, aes(value)) + 
  stat_density(geom="path", position="identity", alpha=0.7) +
  facet_wrap(~ variable, scale="free", ncol=2)
```

@@

```{r figure_S8, fig.cap="Posteriors from the MCMC estimate of the Allen model", fig.width=6, fig.height=6}
ggplot(allen_posteriors, aes(value)) + 
  stat_density(geom="path", position="identity", alpha=0.7) +
  facet_wrap(~ variable, scale="free", ncol=2)
```

@@

>>>>

keywords:
  - Bayesian
  - Structural Uncertainty
  - Nonparametric 
  - Optimal Control
  - Decision Theory
  - Gaussian Processes
  - Fisheries Management

bibliography: components/references.bib
csl: components/ecology.csl
template: components/elsarticle.latex
documentclass: components/elsarticle

## rmarkdown render options
output:
  pdf_document:
    keep_tex: true
---

@@

```{r sensitivity-trends}
source("components/sensitivity-trends.Rmd")
```

@@

write.csv(df_post, "components/data/figure1.csv")
```

@@

```{r figure_2, dependson=c("assemble-models", "plot-options"), fig.cap="Graph of the inferred Gaussian process compared to the true process and maximum-likelihood estimated process.  Graph shows the expected value for the function $f$ under each model.  Two standard deviations from the estimated Gaussian process covariance with (light grey) and without (darker grey) measurement error are also shown.  The training data is also shown as black points.  The GP is conditioned on (0,0), shown as a pseudo-data point.", fig.width=8, fig.height=6}
x_grid_short <- x_grid[1:40]
gp_short <- gp_predict(gp, x_grid_short, burnin=1e4, thin=300)
models_posteriors <- 
  melt(lapply(sample(100, 50), 
              function(i){
    sample_gp <- mvrnorm(1, 
                            gp_short$Ef_posterior[,i],         
                            gp_short$Cf_posterior[[i]])
    data.frame(stock = x_grid_short, 
               GP = sample_gp,
               y = sample_gp,
               ymin = sample_gp - 2 * sqrt(gp_short$E_Vf), 
               ymax = sample_gp + 2 * sqrt(gp_short$E_Vf), 
               True = sapply(x_grid_short,f,0, p),  
               MLE = sapply(x_grid_short,f,0, est$p), 
               Allen = sapply(x_grid_short, allen_f, 0, pardist[i,]), 
               Ricker = sapply(x_grid_short, ricker_f, 0, ricker_pardist[i,]), 
               Myers = sapply(x_grid_short, myers_f, 0, myers_pardist[i,]))
             }), 
       id=c("stock", "y", "ymin", "ymax"))
ggplot(models_posteriors) + 
    geom_ribbon(aes(x=stock, y=y, ymin=ymin, ymax=ymax, group=L1), 
                  fill = "gray80", 
                  data=subset(models_posteriors, variable == "GP")) + 
    geom_line(aes(stock, value, col = variable, 
                  group=interaction(L1,variable)), 
              alpha=.2) + 
    geom_point(data = obs, aes(x,y), alpha = 0.8) + 
    xlab(expression(X[t])) + ylab(expression(X[t+1])) +
    facet_wrap(~variable) + 
    scale_colour_manual(values=colorkey) +  
    theme(legend.position="none")

write.csv(models_posteriors, "components/data/figure2.csv")
write.csv(obs, "components/data/obs.csv")
```

@@

```{r figure_3, fig.width=8, fig.height=6, fig.cap="Outside the range of the training data (Figure 1), the true dynamics (black dots) fall outside the uncertainty (two standard deviations, colored bands) of the structurally incorrect parametric models (Ricker, Myers), but inside the uncertainty predicted by the GP. Points show the stock size simulated by the true model.  Overlay shows the range of states predicted by each model, based on the state observed in the previous time step. The Ricker model always (wrongly) predicts positive population growth, while the actual population shrinks in each step as the initial condition falls below the Allee threshold of the underlying model (Allen).  Note that because it does not assume a parametric form but instead relies more directly on the data, the GP is both more pessimistic and more uncertain about the future state than the parametric models.", dependson="plot-options"}
y <- numeric(8)
y[1] <- 4.5
for(t in 1:(length(y)-1))
      y[t+1] = z_g() * f(y[t], h=0, p=p)
# predicts means, does not reflect uncertainty estimate!
crash_data <- step_ahead_posteriors(y)
crash_data <- subset(crash_data, variable %in% c("GP", "Allen", "Ricker", "Myers"))
ggplot(crash_data) + 
  geom_boxplot(aes(as.factor(as.integer(time)), value, 
                   fill = variable, col=variable), 
               alpha=.7, outlier.size=1, position="identity") + 
#  geom_line(aes(time, value, col = variable, 
#            group=interaction(L1,variable)), alpha=.1) + 
  geom_point(aes(time, stock), size = 3) + 
  scale_fill_manual(values=colorkey[c("GP", "Allen", "Ricker", "Myers")], 
                      guide = guide_legend(override.aes = list(alpha = 1))) +  
  scale_colour_manual(values=colorkey[c("GP", "Allen", "Ricker", "Myers")], 
                      guide = guide_legend(override.aes = list(alpha = 1))) +  
  facet_wrap(~variable) + 
  theme(legend.position="none") + xlab("time") + ylab("stock size")

write.csv(crash_data, "components/data/figure3.csv")
```

@@

ggplot(policies, aes(stock, stock - value, color=method)) +
  geom_line(lwd=1.2, alpha=0.8) + xlab("stock size") + ylab("escapement")  +
  scale_colour_manual(values=colorkey)

write.csv(policies, "components/data/figure4.csv")
```

@@

write.csv(sims_data, "components/data/figure5.csv")
```

@@

# ggplot(actual_over_optimal, aes(value)) + geom_histogram(aes(fill=variable), binwidth=0.1) + 
#  xlab("Total profit by replicate")+ scale_fill_manual(values=colorkey)
# ggplot(actual_over_optimal, aes(value, fill=variable, color=variable)) + # density plots fail when delta fn
#  stat_density(aes(y=..density..), position="stack", adjust=3, alpha=.9) + 
#  xlab("Total profit by replicate")+ scale_fill_manual(values=colorkey)+ scale_color_manual(values=colorkey)

write.csv(actual_over_optimal, "components/data/figure6.csv")
```

@@

(following the Ahrenius relationship, see @Gardiner2009 for the derivation).
As a result, the impact of using a model which underestimates the risk of harvesting
past the critical point is much more severe, since this such a situation
occurs more often.  Conversely, with large enough distance between the
optimal escapement and unstable points relative to the noise, the chance
of a transition becomes vanishingly small and all models can be modeled
near-optimally. Models that underestimate the cost incurred by stock
sizes fluctuating significantly below the optimal escapement level will
not perform poorly as long as those fluctuations are sufficently rare.  Figure 7
shows that the GPDP is little effected by increasing noise or increasing Allee
effects over much of the range.

@@

```{r figure_7, fig.height=6, dependson=c("sensitivity-trends"), fig.cap="The effect of increasing noise or decreasing Allee threshold levels on the net present value of the fishery when managed under the GPDP, relative to managing under the true model (with known parameters).  Other than the focal parameter (noise, Allee threshold), other parameters are held fixed as above to illustrate this effect."}

plot_sigmas <- ggplot(vary_sigma, aes(noise, value)) + 
  geom_point() + ylab("Net present value") + 
  xlab("Level of growth stochasticity (sigma)") + theme_bw() 
plot_allees <- ggplot(vary_allee, aes(pars.C, value)) + 
  geom_point()+ ylab("Net present value") + 
  xlab("Allee threshold stock size") + theme_bw()
multiplot(plot_sigmas, plot_allees, cols=2)

write.csv(vary_sigma, "components/data/vary_sigma.csv")
write.csv(vary_allee, "components/data/vary_sigma.csv")
```

@@

```{r figure_S4, fig.cap="Posteriors from the MCMC estimate of the Ricker model", fig.width=6, fig.height=4}
ggplot(ricker_posteriors, aes(value)) + 
  stat_density(geom="path", position="identity", alpha=0.7) +
  facet_wrap(~ variable, scale="free", ncol=2)
write.csv(ricker_posteriors, "components/data/ricker_posteriors.csv")
```

@@

```{r figure_S6, fig.cap="Posterior distributions from the MCMC estimates of the Myers model", fig.width=6, fig.height=6}
ggplot(myers_posteriors, aes(value)) + 
  stat_density(geom="path", position="identity", alpha=0.7) +
  facet_wrap(~ variable, scale="free", ncol=2)
write.csv(myers_posteriors, "components/data/myers_posteriors.csv")
```

@@

```{r figure_S8, fig.cap="Posteriors from the MCMC estimate of the Allen model", fig.width=6, fig.height=6}
ggplot(allen_posteriors, aes(value)) + 
  stat_density(geom="path", position="identity", alpha=0.7) +
  facet_wrap(~ variable, scale="free", ncol=2)
write.csv(allen_posteriors, "components/data/allen_posteriors.csv")
```

@@

```{r figure_S9, fig.height=6, dependson=c("sensitivity-calc", "export-data"), fig.cap="Sensitivity Analysis.  Histogram shows the ratio of the realized net present value derived when managing under the GPDP over the optimal value given the true model and true parameters. Values near 1 indicate nearly optimal performance. Simulations cover a range of underlying parameter, model, replicate, and noise levels."}
ggplot(sensitivity_dat) + 
  geom_histogram(aes(value, fill=simulation)) + 
  xlim(0,1.0) + 
  theme_bw() + 
  xlab("value as fraction of the optimal") + 
  facet_grid(noise~model)
write.csv(sensitivity_dat, "components/data/sensitivity_dat.csv")
